Annotation-Based Empirical Performance Tuning Using Orio

### 2009-02-12 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@430 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

In this section, we discuss the performance results of several experiments on
a multicore Intel Xeon workstation and a Blue Gene/P (both at Argonne). The
workstation has dual quad-core E5462 Xeon processors (8 cores total) running
at 2.8 GHz (1600 MHz FSB) with 32 KB L1 cache, 12 MB of L2 cache (6 MB shared
per core pair), and 2 GB of DDR2 FBDIMM RAM, running Linux kernel version
2.6.25 (x86-64). Each compute node of the Blue Gene/P is equipped with four
850MHz IBM PowerPC 450 processors with a dual floating-point unit and 2GB
total memory per node, private L1 (32 KB) and L2 (4 MB) caches, a shared L3
cache (8 MB), and running a proprietary lightweight operating system. We used
version 10.1 of the Intel and version 9.0 of the IBM XL C/C++ V9.0 compilers
on the Xeon and Blue Gene/P, respectively. Because of space considerations,
here we discuss a limited number of computational kernels; more performance
results for different computations are available at the Orio project
website~\cite{OrioURL}.

@@

\caption{(a) Compressed sparse row (CSR) format (b) Basic implementation of CSR-based SpMV.}
\label{fig:spmv}
\end{figure}

@@

The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once
and element reuse is only possible for $x$ and $y$. Thus, to optimize SpMV,
one should use compact data structures for $A$ and try to maximize temporal
reuse of $x$ and $y$. One of the most commonly used data structures for
storing a sparse matrix is compressed sparse row (CSR) storage
~\cite{vuduc-thesis},
illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
packed together in a dense array, \texttt{Aval}, and a corresponding array of
integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
designates where each sparse row begins in \texttt{Aval} and
\texttt{Aind}. An implementation of SpMV using CSR storage is
shown in Figure~\ref{fig:spmv}(b).

PETSc's implementation of SpMV used in this experiment exploits matrix
structure by using \textit{inodes}, which represent rows with identical
nonzero structure.  PETSc detects inodes automatically in arbitrary sparse
matrices. For example, the sparse matrix shown in Figure~\ref{fig:spmv}(a)
contains three inodes. The first inode is of size two because it holds two
adjacent rows (i.e., the first and the second rows) with the same nonzero
structure. The second inode contains three consecutive rows with identical
nonzero structure, while the last inode only has a single row. Knowing the
properties of each inode at runtime enables maximum reuse of vector $x$ since
multiple elements of $x$ can be loaded and used exactly once for all rows in
the same inode structure. To accomplish this, one must use a
register-blocking transformation. To optimize the inode-based SpMV routine,
we incorporated a new transformation module inside Orio that implements
various optimization strategies including register blocking, SIMDization,
memory alignment optimization, loop-control optimization, accumulator
expansion, and thread-level parallelization (with OpenMP). We then used Orio
to automatically select the best version of the optimized inode SpMV. Only
the outer loop which iterates over inodes was parallelized by using OpenMP.

@@

Figure~\ref{fig:ex27-cookie-results} shows the performance results on the
Intel Xeon. The problem size is the number of grid points in the $x$ and $y$
directions, not the size of the sparse input matrix. The matrix dimension is
based the grid size and the number of field components computed for each grid
point, for example, an $8 \times 8$ problem involves sparse matrices of
dimension 900 with 17,040 nonzero entries. In this experiment, we ran in
three different multicore modes: SMP (one MPI process, eight
threads/process), dual (four MPI processes, two threads/process), and virtual
node or VN (eight MPI processes, one thread/process). The code tuned by Orio
consistently outperforms the alternatives. Furthermore, the performance of
the hand-tuned code is almost equivalent to that of the icc-optimized simple
code. Performance differences between the Orio version and the hand-tuned
version become more significant as more threads per process are employed to
facilitate OpenMP parallelization. The SMP and VN results show that
optimizations using loop-level parallelism (through OpenMP) achieve much
better performance than using MPI coarse-grained parallelism.
%Moreover, for the multithread cases, tuning using Orio determined that
%the performance benefit of SIMDization is more significant than that of
%thread-level parallelization only when the problem sizes are small. The
%reason for this is due to the overhead introduced by OpenMP from coordinating
%its threads across processors.
%So, each optimized
%code for the many-thread cases comprises two distinct tuned codes to handle

@@

The Blue Gene/P results in Figure~\ref{fig:ex27-bgp-results1} again show
%Blue Gene/P was done for a varied number of nodes (i.e., 1, 8, and 32)
%and the same three different parallel modes as before (i.e., SMP,
%dual, and VN). 
that empirical optimization using Orio produces the best performance for
all cases. 
%outperforming both the hand-tuned and the xlc-optimized codes. 
On this architecture, the performance gap between the xlc-optimized and the
hand-tuned codes is now larger than in the Intel experiment. Similarly, by
exploiting thread-level parallelism, the Orio-optimized code performs better
than the hand-tuned version when the number of threads per process increases
and the number of nodes decreases. For the single-thread cases (VN mode),
thread-level parallelism is not exploited; nevertheless, the Orio version
still performs better than the other two alternatives.
%This validates the
%effectiveness of Orio in empirically tuning sequential codes.

@@

The performance of the sequential 2-D FDTD code for $tmax=500$ and $nx=ny$ is
shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
alone performs better than Pluto for small problem sizes since all input
arrays fit in the L2 cache (insufficient computation to offset Pluto's tiling
overhead). As the array sizes increase, the lack of data reuse impairs the
base code's performance, whereas the Pluto performance remains about the
same.
%We used Orio to generate two code variants: one for small problem sizes
%and another for large problem sizes. 
When the input arrays are small, Orio discovers that applying
Pluto's polyhedral transformations is not beneficial, and therefore,
it employs only its syntactic transformations on the original FDTD
code. For large problem sizes, Orio exploits some of the Pluto's code
transformations and enhances it further with its syntactic
optimizations, resulting in performance consistently and significantly
higher than both the base and Pluto codes (up to 86\% over Pluto).

@@

Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance obtained
for $tmax=500$ and $nx=ny=2000$. The results indicate that whereas icc is
unable to auto-parallelize the code, Pluto detects the existence of pipelined
parallelism and then successfully parallelizes the code. Optimizing and
tuning by Orio additionally improves the Pluto performance by up to
78\%. Moreover, we observe that because of memory contention between the two
quad-core Intel processors, the speedup of both the Pluto and Orio codes
slightly decreases when the number of cores used is greater than four.

@@

\subsubsection{3-D Gauss-Seidel Successive Over-Relaxation Method}  
Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
which is sometimes is referred to as
\emph{successive displacement method}, indicating the dependence of the 
iterations on the ordering. If the ordering is changed, the components
of the new iterations will change as well.
%
%\begin{figure}[ht]
\begin{figure}%[ht]
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for (t=0; t<=T-1; t++) 
  for (i=1; i<=N-2; i++) 
    for (j=1; j<=N-2; j++) 
      A[i][j] = (A[i-1][j-1] + A[i-1][j] 
          + A[i-1][j+1] + A[i][j-1] + A[i][j] 
          + A[i][j+1] + A[i+1][j-1] + A[i+1][j] 
          + A[i+1][j+1]) / 9.0; 
\end{verbatim} 
\end{minipage} 
\end{center}
\caption{3-D Gauss-Seidel code} 
\label{fig:seidel-code} 
\end{figure}
%

>>>>

In this section, we discuss the performance results of several experiments on
a multicore Intel Xeon workstation and a Blue Gene/P (both at Argonne). The
workstation has dual quad-core E5462 Xeon processors (8 cores total) running
at 2.8 GHz (1600 MHz FSB) with 32 KB L1 cache, 12 MB of L2 cache (6 MB shared
per core pair), and 2 GB of DDR2 FBDIMM RAM, running Linux kernel version
2.6.25 (x86-64). Each compute node of the Blue Gene/P is equipped with four
850 MHz IBM PowerPC 450 processors with a dual floating-point unit and 2 GB
total memory per node, private L1 (32 KB) and L2 (4 MB) caches, a shared L3
cache (8 MB), and running a proprietary lightweight operating system. We used
version 10.1 of the Intel and version 9.0 of the IBM XL C/C++ V9.0 compilers
on the Xeon and Blue Gene/P, respectively. Because of space considerations,
here we discuss a limited number of computational kernels; more performance
results for different computations are available at the Orio project
website~\cite{OrioURL}.

@@

\caption{(a) Compressed sparse row (CSR) format; (b) Basic implementation of CSR-based SpMV.}
\label{fig:spmv}
\end{figure}

@@

The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once
and element reuse is possible for $x$ and $y$. Thus, to optimize SpMV,
one should use compact data structures for $A$ and try to maximize temporal
reuse of $x$ and $y$. One of the most commonly used data structures for
storing a sparse matrix is compressed sparse row (CSR) storage
~\cite{vuduc-thesis},
illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
packed together in a dense array, \texttt{Aval}, and a corresponding array of
integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
designates where each sparse row begins in \texttt{Aval} and
\texttt{Aind}. An implementation of SpMV using CSR storage is
shown in Figure~\ref{fig:spmv}(b).

PETSc's implementation of SpMV used in this experiment exploits matrix
structure by using \textit{inodes}, which represent rows with identical
nonzero structure.  PETSc detects inodes automatically in arbitrary sparse
matrices. For example, the sparse matrix shown in Figure~\ref{fig:spmv}(a)
contains three inodes. The first inode is of size two because it holds two
adjacent rows (i.e., the first and the second rows) with the same nonzero
structure. The second inode contains three consecutive rows with identical
nonzero structure, and the last inode only has a single row. Knowing the
properties of each inode at runtime enables maximum reuse of vector $x$ since
multiple elements of $x$ can be loaded and used exactly once for all rows in
the same inode structure. To accomplish this, one must use a
register-blocking transformation. To optimize the inode-based SpMV routine,
we incorporated a new transformation module inside Orio that implements
various optimization strategies including register blocking, SIMDization,
memory alignment optimization, loop-control optimization, accumulator
expansion, and thread-level parallelization (with OpenMP). We then used Orio
to automatically select the best version of the optimized inode SpMV. Only
the outer loop that iterates over inodes was parallelized by using OpenMP.

@@

Figure~\ref{fig:ex27-cookie-results} shows the performance results on the
Intel Xeon. The problem size is the number of grid points in the $x$ and $y$
directions, not the size of the sparse input matrix. The matrix dimension is
based on the grid size and the number of field components computed for each grid
point; for example, an $8 \times 8$ problem involves sparse matrices of
dimension 900 with 17,040 nonzero entries. In this experiment, we ran in
three multicore modes: SMP (one MPI process, eight
threads/process), dual (four MPI processes, two threads/process), and virtual
node or VN (eight MPI processes, one thread/process). The code tuned by Orio
consistently outperforms the alternatives. Furthermore, the performance of
the hand-tuned code is almost equivalent to that of the icc-optimized simple
code. Performance differences between the Orio version and the hand-tuned
version become more significant as more threads per process are employed to
facilitate OpenMP parallelization. The SMP and VN results show that
optimizations using loop-level parallelism (through OpenMP) achieve much
better performance than using MPI coarse-grained parallelism.
%Moreover, for the multithread cases, tuning using Orio determined that
%the performance benefit of SIMDization is more significant than that of
%thread-level parallelization only when the problem sizes are small. The
%reason for this is due to the overhead introduced by OpenMP from coordinating
%its threads across processors.
%So, each optimized
%code for the many-thread cases comprises two distinct tuned codes to handle

@@

The Blue Gene/P results in Figure~\ref{fig:ex27-bgp-results1} again show
%Blue Gene/P was done for a varied number of nodes (i.e., 1, 8, and 32)
%and the same three different parallel modes as before (i.e., SMP,
%dual, and VN). 
that empirical optimization using Orio produces the best performance for
all cases. 
%outperforming both the hand-tuned and the xlc-optimized codes. 
On this architecture, the performance gap between the xlc-optimized and the
hand-tuned codes is now larger than in the Intel experiment. Similarly, by
exploiting thread-level parallelism, the Orio-optimized code performs better
than the hand-tuned version when the number of threads per process increases
and the number of nodes decreases. For the single-thread cases (VN mode),
thread-level parallelism is not exploited; nevertheless, the Orio version
still performs better than the other two choices.
%This validates the
%effectiveness of Orio in empirically tuning sequential codes.

@@

The performance of the sequential 2-D FDTD code for $tmax=500$ and $nx=ny$ is
shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
alone performs better than Pluto for small problem sizes since all input
arrays fit in the L2 cache (insufficient computation to offset Pluto's tiling
overhead). As the array sizes increase, the lack of data reuse impairs the
base code's performance, whereas the Pluto performance remains about the
same.
%We used Orio to generate two code variants: one for small problem sizes
%and another for large problem sizes. 
When the input arrays are small, Orio discovers that applying
Pluto's polyhedral transformations is not beneficial, and therefore
it employs only its syntactic transformations on the original FDTD
code. For large problem sizes, Orio exploits some of the Pluto's code
transformations and enhances them further with its syntactic
optimizations, resulting in performance consistently and significantly
higher than both the base and Pluto codes (up to 86\% over Pluto).

@@

Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance
obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
whereas icc is unable to auto-parallelize the code, Pluto detects the
existence of pipelined parallelism and then successfully parallelizes
the code. Locality-exploiting optimizations by Orio additionally
improve the Pluto performance by up to 78\%. Moreover, we observe
that because of memory contention between the two quad-core Intel
processors, the speedup of both the Pluto and Orio codes slightly
decreases when the number of cores used is greater than four.

@@

\subsubsection{3-D Gauss-Seidel Successive Over-Relaxation Method}  
Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
which is sometimes referred to as
\emph{successive displacement method}, indicating the dependence of the 
iterations on the ordering. If the ordering is changed, the components
of the new iterations will change as well.
%
%\begin{figure}[ht]
\begin{figure}%[ht]
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for (t=0; t<=T-1; t++) 
  for (i=1; i<=N-2; i++) 
    for (j=1; j<=N-2; j++) 
      A[i][j] = (A[i-1][j-1] + A[i-1][j] 
          + A[i-1][j+1] + A[i][j-1] + A[i][j] 
          + A[i][j+1] + A[i+1][j-1] + A[i+1][j] 
          + A[i+1][j+1]) / 9.0; 
\end{verbatim} 
\end{minipage} 
\end{center}
\caption{3-D Gauss-Seidel code} 
\label{fig:seidel-code} 
\end{figure}
%
